---
title: "Decathlon PCA"
author: "Carla Cohen"
date: "2024-06-25"
output: html_document
---

# Introduction to R coding

Today we are going to do some coding using the programming language R. 

This document is in R markdown format, which allows you to intersperse lines of code and free text. 

The grey boxes are the code, they always begin "```{r}" and end "```"
To add a new chunk of code, click the green button in the top right hand corner of this window and choose "R". 

This first chunk tells the document any general set up instructions (no need to worry about this today).

Activity 1: Press the "play" button to run this chunk. 
You will see some outputs in the console that tells you these packages are loaded.  
These packages contains the instructions and the data that we will be using.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(FactoMineR)
library(Rfast)
library(factoextra)
```


# Using R markdown 

Activity 2: Add a new code chunk!
Write a simple sum e.g. 1+2
When you press the "play" button (or Ctrl+enter) you will see the output of the code below the chunk i.e. "3"
[1] means the first line
```{r}
1+2
```


# Assigning variables

We can assign values to a variable using the "<-" notation
e.g. 
a <- 5

Now you will see these values appear in the Global Environment (top right pane). 
If you want to know what the variable is, you can print it out

Activity 3: Assign the letters a and b to some numbers

```{r}
a <- 6
b <- 8

```

Activity 4: Print one of your variables.  

```{r}
a
```
If you want to make the output more meaningful, you can add some text using the print command
e.g. 
print ("This is variable a")
a

Activity 5: Print some text and a variable 

```{r}
print ("This is variable a")
a
print ("My favourite number is ")
b
```
Then we can do a sum using our variables

Activity 6: Write a sum using your variables

```{r}
a+b
```

Those variables were whole numbers, or integers. 

# Vectors

We can also use vectors, which is like a list of numbers (although bear in mind that a "list" is also another data structure in R that we aren't going to use today!).

A vector looks like this: 
```{r}
my_vector <- c(1, 2, 3, 4, 5)
my_vector
```

You can also do sums using the vector

Activity 7: Multiply my vector by 2 
Note: use * for multiply


```{r}
my_vector*2
```

# Example dataset

To do more interesting things we will need to have some data. 
We can read in our own data files from our experiment, but in this case we will use one that is already built into R. 

This command produces a variable called "decathlon" in the global environment

```{r}
data(decathlon)

```

Activity 8: Print the variable "decathlon"
Can you describe what this dataset it showing? 
What are the rows?
What are the columns?
What is the format of the data (numbers/text?)  

```{r}
decathlon
```


# Introduction to the decathlon dataset

This dataset contains the results of decathlon events during two athletic meetings which took place one month apart in 2004: the Olympic Games in Athens (23 and 24 August), and the Decastar 2004 (25 and 26 September). 

For both competitions, the following information is available for each athlete: 
- performance for each of the 10 events   
- total number of points (for each event, an athlete earns points based on performance; here the sum of points scored) 
- final ranking  


# Working with dataframes
A table like the decathlon data is called a dataframe in R. 
You can get the information for a particular column using the $ symbol. 
For example using decathlon$Shot.put gives you all the data from the column "Shot.put"

Actvity 9: Get the data from the column of your favourite event 

```{r}
decathlon$`100m`
```
You can see that it just gives you the numbers, you have lost the information related to which competitor it was. 

# Plotting a graph of the data

We can have a look at the distribution of the data, for example, how far did the different competitors throw the javelin?

There are some very easy plotting functions in R (spoiler: they don't look very pretty!)
We can use
plot(data)
hist(data)

Activity 9: Use the plot and hist functions to draw a graph of the javelin data  
When using plot, can you also sort the data to make it more meaningful?  
NB to access the help for a function, type ?function in the console e.g. ?plot 

```{r}
plot(sort(decathlon$Javeline))
```

Question: What do these graphs tell you? 
  
# Finding out more about the javelin event? 

We can quickly calculate some basic statistics of the javelin data, for example: 
What is the mean javelin throw? What is the furthest?

Activity 10: Calculate the mean javelin throw and furthest javelin throw. 
Hint - use mean() and max()  

```{r}
mean(decathlon$Javeline)
max(decathlon$Javeline)

```
Activity 11: Copy your code to plot the sorted javelin data. 
Add a line for mean javelin throw to the plot using abline(h = ....)
Can you change the colour of the line?  
Can you add a title?  

```{r}
plot(sort(decathlon$Javeline))
abline(h = mean(decathlon$Javeline), col = "red")
```


# Are athletes typically good at all throwing events?

Did javelin score correlate with shot put?

Activity 12: Draw a graph of javelin distance vs shot put distance using plot ()
Add a regression line (line of best fit) to the plot
- first calculate the regression line using lm()
- draw the graph using plot()
- add the regression line to the plot using abline()

```{r}
# calculate the regression line using lm()

reg <- lm(Shot.put ~ Javeline, data = decathlon)

# Draw the plot
plot(decathlon$Javeline, decathlon$Shot.put)
abline(reg, col = "blue")

```

You can quantify the correlation between javelin and shot put distance by calculating a correlation score (between 0 and 1) 

Activity 13: Calculate the correlation score between javelin & shot put using cor()
Hint - this will look very similar to the code used for the plot!

```{r}
cor(decathlon$Javeline, decathlon$Shot.put)
cor(decathlon$High.jump, decathlon$Long.jump)
```

If you have time, compare two other events in the same way. Do they correlate better or worse than javelin and shot put? 

```{r}

```

```{r}

```

# What is the variance of each event? 
This represents how widely variable the scores for each event are 
e.g. the range of high jump scores might be much smaller than the range of 1500m scores. 

Activity 14: Calculate the variance for each event using colVars()
Hint you can do this for the whole dataset at once! 

```{r}
colVars(decathlon)
```

# Scale the data

To make more in-depth analysis we can "scale the data". This means that each event will have a mean of 0 and a standard deviation of 1. 
I will explain this in more detail if time! 

The point of doing this is that we can compare across the different events more fairly. 

Activity 15: Scale the data using the following steps.
- remove the column for "Competition" using select () and assign it to a new dataframe called dec. Why do we need to do that?
- scale dec and assign it to a new dataframe called dec_scaled 
This produces a scaled dataset and draws a graph of the scaled javelin data.  

```{r}
# remove the column "Competition" because it isn't numeric
dec <- select(decathlon, -Competition)
dec
# scale the data
dec_scaled <- data.frame(scale(dec))
```

Activity 16: Plot the scaled data for Javelin using hist()

```{r}
hist(dec_scaled$Javeline)
```



# Principal component analysis

We can perform a Principal Component Analysis (PCA) to look at how all of the events go together. 
This analysis produces 3 outputs
1. numerical summary (ignore)
2. Graph of individuals: This shows you how similar the individual competitors were to one another. 
3. Graph of variables: This shows you how much each event had on the overall result. 

Activity 17: Perform a PCA on the scaled decathlon data using PCA()

```{r}
PCA(dec_scaled)
```

Questions:
Which sports should you be good at if you want to win decathlon? 
Which sports do competitors tend to be good at in groups?

# Interpretation

Add your interpretation here

# Plot the PCA of competitors coloured by the competition they took part in

Extended activity: Did the competitors perform similarly in both competitions?

Add back the Competition column to the scaled results

```{r}

```

Re-do the PCA, specifying Competition as a qualitative variable. 
Then plot the PCA of individuals coloured by the competition using fviz_pca_ind() and habillage  
Hint: use repel=TRUE to read the names more easily


```{r}
res.pca <- PCA()

fviz_pca_ind(res.pca, ) 
```





# Save your script
Email me carla.cohen@ndorms.ox.ac.uk if you would like to have a copy of it to keep.  

# Knit the script
This produces a nice html output with all the text, code and plots. 

# Reference
This tutorial was based on an analysis by Clement Ponsonnet
https://github.com/clemonster/decathlon-pca

